Reproducing anaylses of Beninca et al (2008)

### 2016-03-29 code/report.Rmd
revision to attempt to fix behaviour of gather()

@@

Data is in wide format, so change it to long:
```{r}
spp.abund <- gather(spp.abund, key="variable", value="value",
                    3:12)
spp.abund$variable <- as.factor(spp.abund$variable)
str(spp.abund)
```

@@

nuts <- select(nuts, -X, -X.1)
nuts <- nuts[-349:-8163,]
nuts$Date <- dmy(nuts$Date)
nuts <- select(nuts, -NO2, -NO3, -NH4)
nuts <- gather(nuts, "variable", "value", 3:4)
nuts$variable <- as.factor(nuts$variable)
str(nuts)
```

@@

The exact method for identifying the "long sequences of zeros" was not given. However, the transformed data with zeros removed was provided, and from this we can get the day range of the retained data:
```{r}
tr <- read.csv(text=getURL(paste0(repo,"/data/original/transformed_data_Nature2008.csv")),
               skip=7, na.string="")
tr <- tr[,-14:-24] ## remove bad columns
tr <- tr[-693:-694,] ## remove last two rows (contain summary stats)
tr <- gather(tr, key="Species", value="Abundance", 2:13) ## make long format
tr$Species <- as.factor(tr$Species)
levels(tr$Species)[levels(tr$Species)=="Calanoids"] <- "Calanoid.copepods" # rename a level (and next line)
levels(tr$Species)[levels(tr$Species)=="Total.Dissolved.Inorganic.Nitrogen"] <- "Total.dissolved.inorganic.nitrogen"
names(tr)[2] <- "variable" # rename a couple of variables...
names(tr)[1] <- "Day.number"
tr$Day.number <- as.numeric(as.character(tr$Day.number)) # make Day.number numeric
## get day range, in order to use this to constrain later data
min.max.day <- group_by(tr, variable) %>%
  summarise(min.day=min(Day.number[!is.na(Abundance)]), max.day=max(Day.number[!is.na(Abundance)]))
```

@@

Do the time-delayed embedding, and save the data for the graphs:
```{r, message=FALSE, error=TRUE}
all.species <- unique(as.character(led$variable))
diverg <- matrix(NA, time.steps, length(all.species))
colnames(diverg) <- all.species
for(i in 1:length(all.species)) {
  print(all.species[i])
  tr.fs <- filter(final, variable==all.species[i])$y
  diverg[,i] <- as.numeric(try(lyap_k(tr.fs,
                                      m=embedding.dimension,
                                      d=time.delay,
                                      k=10, # number of considered neighbours 20
                                      ref=40, # number of points to take into account 100
                                      t=Theiler.window,
                                      s=time.steps,
                                      eps=10 # radius where to find nearest neighbours 10
                                      )))
}
## a bit of a fudge with the translation to days
diverg <- as.data.frame(cbind(days=1:time.steps, diverg))
diverg <- gather(diverg, Species, Difference, 2:10)
diverg$days <- diverg$days*3.35
```

@@

Perform the interpolation:
```{r}
all.data <- na.omit(all.data)
mt <- plyr::dlply(all.data,
                  "variable",
                  function(xx) pracma::interp1(x=xx$Day.number,
                                               y=xx$value,
                                               xi=xout,
                                               method="cubic"))
## Aside: the duplicated date that was previously fixed was only discovered by a warning message
## given by the pracma::interp1 function!
mt <- as.data.frame(mt)
mt <- cbind(Day.number=xout, mt)
mt <- gather(mt, variable, value, 2:13)
```

@@

from.steve <- gather(from.steve, Species, Abundance, 2:13)
names(from.steve) <- c("Day.number", "variable", "value")
from.steve$value_steve <- from.steve$value 
from.steve <- select(from.steve, -value)

@@

cors <- data.frame(dist=1:12, rsq_vals)
cors.long <- gather(cors, key=dist)
names(cors.long) <- c("Prediction_distance", "Variable", "Correlation")
cors.long$Variable <- as.factor(cors.long$Variable)

@@

## read in the data of the original article that came direct from Elisa
original_preds <- read.csv(text=getURL(paste0(repo,"/data/original/original_rsquared.csv")), skip=0, header=T)
opw <- gather(original_preds, Species, r_squared, 2:25)
opw <- separate(opw, Species, c("Species", "Model"))
opw$Species <- as.factor(opw$Species)
opw$Model <- as.factor(opw$Model)
name.mapping <- data.frame(Species=c("cyclo", "roti", "cala", "pico", "nano",
                                   "dia", "P", "N", "bact", "ostra", "harpa",
                                   "proto"),
                           Variable=c("Cyclopoids", "Rotifers", "Calanoid.copepods",
                                  "Picophytoplankton", "Nanophytoplankton",
                                  "Filamentous.diatoms", "Soluble.reactive.phosphorus",
                                  "Total.dissolved.inorganic.nitrogen", "Bacteria",
                                  "Ostracods", "Harpacticoids", "Protozoa"))
opw <- full_join(opw, name.mapping)

>>>>

# R options

```{r}
options(stringsAsFactors=FALSE)
```

@@

Data is in wide format, so change it to long:
```{r}
spp.abund <- gather(spp.abund, key="variable", value="value",
                    3:12, convert=FALSE)
#spp.abund$variable <- as.factor(spp.abund$variable)
str(spp.abund)
```

@@

nuts <- select(nuts, -X, -X.1)
nuts <- nuts[-349:-8163,]
nuts$Date <- dmy(nuts$Date)
nuts <- select(nuts, -NO2, -NO3, -NH4)
nuts <- gather(nuts, "variable", "value", 3:4, convert=F)
#nuts$variable <- as.factor(nuts$variable)
str(nuts)
```

@@

The exact method for identifying the "long sequences of zeros" was not given. However, the transformed data with zeros removed was provided, and from this we can get the day range of the retained data:
```{r}
tr <- read.csv(text=getURL(paste0(repo,"/data/original/transformed_data_Nature2008.csv")),
               skip=7, na.string="")
tr <- tr[,-14:-24] ## remove bad columns
tr <- tr[-693:-694,] ## remove last two rows (contain summary stats)
tr <- gather(tr, key="Species", value="Abundance", 2:13, convert=F) ## make long format
#tr$Species <- as.factor(tr$Species)
levels(tr$Species)[levels(tr$Species)=="Calanoids"] <- "Calanoid.copepods" # rename a level (and next line)
levels(tr$Species)[levels(tr$Species)=="Total.Dissolved.Inorganic.Nitrogen"] <- "Total.dissolved.inorganic.nitrogen"
names(tr)[2] <- "variable" # rename a couple of variables...
names(tr)[1] <- "Day.number"
tr$Day.number <- as.numeric(as.character(tr$Day.number)) # make Day.number numeric
## get day range, in order to use this to constrain later data
min.max.day <- group_by(tr, variable) %>%
  summarise(min.day=min(Day.number[!is.na(Abundance)]), max.day=max(Day.number[!is.na(Abundance)]))
```

@@

Do the time-delayed embedding, and save the data for the graphs:
```{r, message=FALSE, error=TRUE}
all.species <- unique(as.character(led$variable))
diverg <- matrix(NA, time.steps, length(all.species))
colnames(diverg) <- all.species
for(i in 1:length(all.species)) {
  print(all.species[i])
  tr.fs <- filter(final, variable==all.species[i])$y
  diverg[,i] <- as.numeric(try(lyap_k(tr.fs,
                                      m=embedding.dimension,
                                      d=time.delay,
                                      k=10, # number of considered neighbours 20
                                      ref=40, # number of points to take into account 100
                                      t=Theiler.window,
                                      s=time.steps,
                                      eps=10 # radius where to find nearest neighbours 10
                                      )))
}
## a bit of a fudge with the translation to days
diverg <- as.data.frame(cbind(days=1:time.steps, diverg))
diverg <- gather(diverg, Species, Difference, 2:10, convert=F)
diverg$days <- diverg$days*3.35
```

@@

Perform the interpolation:
```{r}
all.data <- na.omit(all.data)
mt <- plyr::dlply(all.data,
                  "variable",
                  function(xx) pracma::interp1(x=xx$Day.number,
                                               y=xx$value,
                                               xi=xout,
                                               method="cubic"))
## Aside: the duplicated date that was previously fixed was only discovered by a warning message
## given by the pracma::interp1 function!
mt <- as.data.frame(mt)
mt <- cbind(Day.number=xout, mt)
mt <- gather(mt, variable, value, 2:13, convert=F)
```

@@

from.steve <- gather(from.steve, Species, Abundance, 2:13, convert=F)
names(from.steve) <- c("Day.number", "variable", "value")
from.steve$value_steve <- from.steve$value 
from.steve <- select(from.steve, -value)

@@

cors <- data.frame(dist=1:12, rsq_vals)
cors.long <- gather(cors, key=dist, convert=F)
names(cors.long) <- c("Prediction_distance", "Variable", "Correlation")
#cors.long$Variable <- as.factor(cors.long$Variable)

@@

## read in the data of the original article that came direct from Elisa
original_preds <- read.csv(text=getURL(paste0(repo,"/data/original/original_rsquared.csv")), skip=0, header=T)
opw <- gather(original_preds, Species, r_squared, 2:25, convert=F)
opw <- separate(opw, Species, c("Species", "Model"))
#opw$Species <- as.factor(opw$Species)
#opw$Model <- as.factor(opw$Model)
name.mapping <- data.frame(Species=c("cyclo", "roti", "cala", "pico", "nano",
                                   "dia", "P", "N", "bact", "ostra", "harpa",
                                   "proto"),
                           Variable=c("Cyclopoids", "Rotifers", "Calanoid.copepods",
                                  "Picophytoplankton", "Nanophytoplankton",
                                  "Filamentous.diatoms", "Soluble.reactive.phosphorus",
                                  "Total.dissolved.inorganic.nitrogen", "Bacteria",
                                  "Ostracods", "Harpacticoids", "Protozoa"))
opw <- full_join(opw, name.mapping)
